 Allocation scoring rules as a generalization of quantile scoring rules

### 2023-07-10 alloscore_manuscript/allocation.tex
Update comment colors latex, and added an example allocation file that works on my comp
fix iter_pic, start allocation score discussion for paper intro

\usepackage{soul}
\usepackage{color}
\def\elr#1{{\color{cyan}\textbf{ELR:[#1]}}}
\def\apg#1{{\color{red}\textbf{APG:[#1]}}}
\def\bwr#1{{\color{violet}\textbf{BWR:[#1]}}}

@@

>>>>

\usepackage{soul}
\usepackage{xcolor}
\def\elr#1{{\color{cyan}\textbf{ELR:[#1]}}}
\def\apg#1{{\color{red}\textbf{APG:[#1]}}}
\def\bwr#1{{\color{violet}\textbf{BWR:[#1]}}}

@@

\subsection{Allocation Scoring}

In a nutshell, the allocation score of a multivariate forecast $F$ is the realized unmet demand when $F$ is used to allocate so as to minimize expected unmet 
need --- that is, when we distribute a limited amount $K$ of our resource so that total expected unmet need according to $F$,
\begin{align}
\sum_i E_{F_i}[\text{unmet need in location $i$}]
\end{align}
is minimized, where the $F_i$ are the marginal distributions of $F$.
This stochastic allocation problem has an intuitively appealing solution: allocate so that the probabilities of need exceeding allocation 
in various locations are as close to each other as possible (see methods).  In general, this will lead to the allocations provided by $F$ being either zero or the quantiles of $F_i$ for some \emph{single} probability level $\tau$.

For example, suppose we have a forecast $F$ for need in two locations with $F_1 = \mathrm{Unif}(0,4)$ and $F_1 = \mathrm{Unif}(4,8)$. If we have $K = 10$ units
of our resource available, the optimal allocation according to $F$ would be 3 units in location 1 and 7 units in location 2. Then both locations would be allocated at the quantiles for probability level $\tau = .75$.  If, however, we only have $K = 3$ units available, all 3 would go to location 2 since it is not
possible to match probabilities of need exceeding allocation given our resource constraint.

### 2023-07-10 notes/allocation_notes.rmd
fix iter_pic, start allocation score discussion for paper intro

Let $F,F_i$ be general distributions such as forecaster beliefs or reports about $Y$, and $G$ be the true distribution of $Y$. 
For a general scoring rule $S$, we define the $S$-entropy of $F$ as 
\begin{align}
e_S(F) &:= S(F, F) = E_F[S(F, Y)]
\end{align}
and the $S$-entropy of the random variable $Y$ as
\begin{align}
e_S(Y) &:= e_S(G).
\end{align}
Applying this to $S_\alpha$, we can define the $g,\alpha$-entropy of $F$ as
\begin{align}
e_{\alpha}(F) &:= S_{\alpha}(F, F) = E_F[s_{\alpha}(q_{\alpha,F}, Y)] .
\end{align}
A forecaster believing that $Y \sim F$ will expect to receivq_e a score of $e_{\alpha}(F)$ by reporting their $\alpha$ quantile, while an
omniscient forecaster \emph{knowing} that $Y\sim G$ will expect a score of $e_S(Y)$ for reporting $q_{\alpha,Y}$. Note that if 
$F_{y_0} = \mathbb{1}\{x \geq y_0\}$ is deterministic, then 
\begin{align}
e_{\alpha}(F_{y_0}) = E_{F_{y_0}}[s_{\alpha}(q_{\alpha, F_{y_0}}, Y)] = s_{\alpha}(y_0, y_0) = 0. \label{eqn:entropy_determ}
\end{align}
Similarly, if $F_{\mathbf{y}_n} = \mathrm{Unif}\{y_1, \ldots, y_n\}$ is the empirical distribution of a sample $\mathbf{y}_n = \{y_1, \ldots, y_n\}$, then
\begin{align}
e_{\alpha}(F_{\mathbf{y}_n}) = E_{F_{\mathbf{y}_n}}[s_{\alpha}(q_{\alpha, F_{\mathbf{y}_n}}, Y)] 
& = \frac{1}{n}\sum_{i=1}^{n} s_{\alpha}(y_{\lceil \alpha n \rceil}, y_i) \\
& = \frac{1}{n} \left\{(1-\alpha)\sum_{i=\lceil \alpha n \rceil}^{n}g(y_i) - g(y_{\lceil \alpha n \rceil}) + 
\alpha \sum_{i=1}^{\lceil \alpha n \rceil} g(y_{\lceil \alpha n \rceil}) - g(y_i) \right\}
\end{align}
which we also call the $g,\alpha$-entropy $e_{\alpha}(\mathbf{y}_n)$ of the sample $\mathbf{y}_n$.

@@

The advantage of this dual formulation under separability is that when considering a candidate $\lambda$ for $\lambda^{\star}$,
 we can calculate each summand 
$L_{D,i}(\lambda):=\underset{0 \leq x}{\min} \left\{\lambda w_i x + \sbar_{F,i}(x)\right\}$ independently. 
To this end, let $Z_{\lambda,i}$ be the set of $x \in \mathbb{R}$ where the weighted subgradient of $-\sbar_{F_i}(x)$ contains 
$\lambda$ (c.f. \eqref{scalar_Z_deriv}):
\begin{align}
Z_{\lambda,i} := \left\{x \mid 
\lambda \in -w_i^{-1}\partial\sbar_{F_i}(x) = w_i^{-1}\kappa_i g_i'(x)(\alpha_i - [F_i(x-), F_i(x)])\right\}. \label{dual_comp_deriv}
\end{align}
Then for $\lambda>0$ we have
\begin{align}
L_{D,i}(\lambda) = \min \left\{\lambda w_i x + \sbar_{F_i}(x) \mid x \in \{0\} \cup \{Z_{\lambda,i} \cap (0,\infty)\}  \right\}
\end{align}
Note that 
\begin{itemize}
\item $Z_{\lambda,i}$ will contain any (necessarily closed) interval where $\sbar_{F_i}$ is linear with slope $-\lambda w_i$,
\item unless $\lambda=0=O_i$, we have $\sup Z_{\lambda,i} < \infty$ since in this case $\lambda w_i x + \sbar_{F,i}(x)$ is 
eventually monotonically increasing in $x$,
\item $Z_{\lambda,i}\cap [0,\infty) = \varnothing$ for $\lambda > \max_{x\geq 0}\{w_i^{-1}\kappa_i g_i'(x)(\alpha_i - F_i(x))\}$.
\end{itemize}

@@

## Special Case I: $F_i$ from same location-scale family and $w_i = 1$, $O_i = O, U_i = 1$.

Also assume the constraint is active, i.e., 
\begin{align}
K < \sum F_i^{-1}(\alpha)= \sum \mu_i + \sigma_i\Phi^{-1}(\alpha)
\end{align}
where $\Phi$ is the standard CDF for the family and $\alpha = (1+O)^{-1}$. Then setting
\begin{align}
\sum F_i^{-1}(\alpha - \lambda) = \sum \mu_i + \sigma_i\Phi^{-1}(\alpha - \lambda) = K
\end{align}
we get
\begin{align}
\alpha - \lambda = 
\Phi\left(\frac{K - \sum \mu_j}{\sum \sigma_j} \right):= F(K)
\end{align}

and the Bayes act when the constraint is set to $K$ is 
\begin{align} 
 \nonumber Q_i(K) = F_i^{-1}(F(K)) &= \mu_i +\sigma_i\Phi^{-1}(F(K)) \\ 
 \nonumber &= \mu_i + \sigma_i\left(\frac{K - \sum \mu_j}{\sum \sigma_j} \right) \\
 &= \mu_i + \tilde{\sigma}_i(K - \sum \mu_j)
\end{align}

@@

- $Q_i(K)$ here does not depend on $O$ unlike when the constraint is inactive or the $O_i$ differ.

>>>>

Let $F,F_i$ be general distributions such as forecaster beliefs or reports about $Y$, and $G$ be the true distribution of $Y$. 
For a general scoring rule $S$, we define the $S$-entropy of $F$ as 
\begin{align}
e_S(F) &:= S(F, F) = E_F[S(F, Y)]
\end{align}
and the $S$-entropy of the random variable $Y$ as
\begin{align}
e_S(Y) &:= e_S(G).
\end{align}
Applying this to $S_\alpha$, we can define the $g,\alpha$-entropy of $F$ as
\begin{align}
e_{\alpha}(F) &:= S_{\alpha}(F, F) = E_F[s_{\alpha}(q_{\alpha,F}, Y)] .
\end{align}
A forecaster believing that $Y \sim F$ will expect to receiv`e a score of $e_{\alpha}(F)$ by reporting their $\alpha$ quantile, while an
omniscient forecaster \emph{knowing} that $Y\sim G$ will expect a score of $e_S(Y)$ for reporting $q_{\alpha,Y}$. Note that if 
$F_{y_0} = \mathbb{1}\{x \geq y_0\}$ is deterministic, then 
\begin{align}
e_{\alpha}(F_{y_0}) = E_{F_{y_0}}[s_{\alpha}(q_{\alpha, F_{y_0}}, Y)] = s_{\alpha}(y_0, y_0) = 0. \label{eqn:entropy_determ}
\end{align}
Similarly, if $F_{\mathbf{y}_n} = \mathrm{Unif}\{y_1, \ldots, y_n\}$ is the empirical distribution of a sample $\mathbf{y}_n = \{y_1, \ldots, y_n\}$, then
\begin{align}
e_{\alpha}(F_{\mathbf{y}_n}) = E_{F_{\mathbf{y}_n}}[s_{\alpha}(q_{\alpha, F_{\mathbf{y}_n}}, Y)] 
& = \frac{1}{n}\sum_{i=1}^{n} s_{\alpha}(y_{\lceil \alpha n \rceil}, y_i) \\
& = \frac{1}{n} \left\{(1-\alpha)\sum_{i=\lceil \alpha n \rceil}^{n}g(y_i) - g(y_{\lceil \alpha n \rceil}) + 
\alpha \sum_{i=1}^{\lceil \alpha n \rceil} g(y_{\lceil \alpha n \rceil}) - g(y_i) \right\}
\end{align}
which we also call the $g,\alpha$-entropy $e_{\alpha}(\mathbf{y}_n)$ of the sample $\mathbf{y}_n$.

@@

The advantage of this dual formulation under separability is that when considering a candidate $\lambda$ for $\lambda^{\star}$,
 we can calculate each summand 
$L_{D,i}(\lambda):=\underset{0 \leq x}{\min} \left\{\lambda w_i x + \sbar_{F,i}(x)\right\}$ independently. 
To this end, let $Z_{\lambda,i}$ be the set of critical points in the $i$th coordinate corresponding to the value $\lambda$, that is,
the set
\begin{align}
Z_{\lambda,i} := \left\{x \in \mathbb{R} \cup \infty \mid 
\lambda \in -w_i^{-1}\partial\sbar_{F_i}(x) = w_i^{-1}\kappa_i g_i'(x)(\alpha_i - [F_i(x-), F_i(x)])\right\} \label{dual_comp_deriv}
\end{align}
of $x$ where the weighted subgradient of $-\sbar_{F_i}(x)$ contains 
$\lambda$ (c.f. \eqref{scalar_Z_deriv}). (Allowing $x=\infty$ accommodates the case when $\lambda=0$ and $\alpha_i = 1$.)
Since $\lambda w_i x + \sbar_{F,i}(x) \geq 0$ for $\lambda \geq 0$, we have
\begin{align}
L_{D,i}(\lambda) = \min \left\{\lambda w_i x + \sbar_{F_i}(x) \mid x \in \{0\} \cup \{Z_{\lambda,i} \cap (0,\infty)\}  \right\}
\end{align}
Note that 
\begin{itemize}
\item $Z_{\lambda,i}$ will contain any (necessarily closed) interval where $\sbar_{F_i}$ is linear with slope $-\lambda w_i$,
\item unless $\lambda=0=O_i$, we have $\sup Z_{\lambda,i} < \infty$ since in this case $\lambda w_i x + \sbar_{F,i}(x)$ is 
eventually monotonically increasing in $x$,
\item $Z_{\lambda,i}\cap [0,\infty) = \varnothing$ for $\lambda > \max_{x\geq 0}\{w_i^{-1}\kappa_i g_i'(x)(\alpha_i - F_i(x))\}$, 
and in this case we have $L_{D,i}(\lambda) = \sbar_{F_i}(0)$.
\end{itemize}

@@

## Special Case I: $F_i$ from same location-scale family and $w_i = 1$, $\alpha_i = \alpha$.

Also assume the constraint is active, i.e., 
\begin{align}
K < \sum F_i^{-1}(\alpha)= \sum \mu_i + \sigma_i\Phi^{-1}(\alpha)
\end{align}
where $\Phi$ is the base cdf for the family. Then solving for $\lambda^{\star}$ in
\begin{align}
\sum F_i^{-1}(\alpha - \lambda^{\star}) = \sum \mu_i + \sigma_i\Phi^{-1}(\alpha - \lambda^{\star}) = K
\end{align}
we get
\begin{align}
\alpha - \lambda^{\star} = 
\Phi\left(\frac{K - \sum \mu_j}{\sum \sigma_j} \right):= F(K).
\end{align}
When $\alpha = 1$, we might interpet $F(K)$ as the forecast probability under an independence assumption that $K$ will be sufficient to meet demand.

We can now use the assumption that the $F_i$ have a common base distribution to express the Bayes act as
\begin{align} 
 \nonumber x_i^{F,K} = F_i^{-1}(F(K)) &= \mu_i +\sigma_i\Phi^{-1}(F(K)) \\ 
 \nonumber &= \mu_i + \sigma_i\left(\frac{K - \sum \mu_j}{\sum \sigma_j} \right) \\
 &= \mu_i + \tilde{\sigma}_i(K - \sum \mu_j)
\end{align}

@@

- The $x_i^{F,K}$ here do not depend on $\alpha$ unlike when the constraint is inactive or the $\alpha_i$ differ. Intuitively, when the distibutions $F_i$
all have the same intrinsic shape, a change in $\alpha$ changes the level of marginal expected benefit identically in each coordinate, so that an equality
of these expected benefits is maintained.
